---
title: Data visualization with ggplot2
author: Data Carpentry contributors
minutes: 60
editor_options: 
  chunk_output_type: console
---

------------

> ### Learning Objectives
>
> * Produce scatter plots, boxplots, and time series plots using ggplot.
> * Describe what faceting is and apply faceting in ggplot.
> * Modify the aesthetics of a ggplot plot (including axis labels and color).
> * Build complex and customized plots from data in a data frame.

--------------

**`ggplot2`** is included in the **`tidyverse`** package.

```{r load-package, message=FALSE, purl=FALSE}
library(tidyverse)
```

load the data from prev. lesson

```{r load-data, eval=FALSE, purl=FALSE}
surveys_complete <- read_csv("data_output/surveys_complete.csv")
```

look at our data!

```{r}
head(surveys_complete)
```

## Plotting with **`ggplot2`**

build a plot using programming commands

- reproducible
- documented
- easy to change underlying data
- easy to change visualisation type
- reusable
- helper libraries


**`ggplot2`** functions like data in the 'long' format, i.e., a column for every dimension,
and a row for every observation. Well-structured data will save you lots of time
when making figures with **`ggplot2`**


template:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

indicate the data to use

```{r, eval=FALSE, purl=FALSE}
ggplot(data = surveys_complete)
```

map variables to axes and colors using the aesthetic (`aes`) function

```{r, eval=FALSE, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length))
```

add 'geoms' – graphical representations of the data

many available

use the `+` operator

`+` goes at the end of the line (not begining of next line)

`+` vs `%>%` 

```{r first-ggplot, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

<!-- Challenge -->
<!-- Make a scatter (x-y) plot that shows weight measurements collected every year but only for data collected after 1990. Use both the `%>%` and `+` symbols in your answer -->
<!-- hint: remember the function `filter()` -->

<!-- ```{r} -->
<!-- surveys_complete_post1990 <- surveys_complete %>% filter(year > 1990) -->
<!-- ggplot(data = surveys_complete_post1990, mapping = aes(x = year, y = weight)) + -->
<!--   geom_point() -->

<!-- ``` -->

data investigation time!

start by making a plot like we did 
then modify it to display more information *or* less information more clearly

overplotting:  add transparency (`alpha`)

```{r adding-transparency, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.4)
```

are species different here? -> add color

```{r color-by-species-1, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.4, aes(color = species_id))
```


> ### Challenge - scatter plot

> ### Challenge: hex plot - optional  


## Boxplot

visualize the distribution of weight within each species:

(all we need to do is change the geom!)

all the built-in geoms: https://ggplot2.tidyverse.org/reference/

```{r boxplot, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
    geom_boxplot()
```

hide less: show points

```{r boxplot-with-points, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
    geom_boxplot() +
    geom_jitter(height = 0, alpha = 0.3)
```

layer order

```{r boxplot-with-points, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight)) +
    geom_jitter(height = 0, alpha = 0.3)+
    geom_boxplot() 
```

color apply to all layers -> put in top `ggplot(aes())`

```{r color-by-species-2, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = species_id, y = weight, color = species_id)) +
    geom_jitter(height = 0, alpha = 0.3)+
    geom_boxplot() 
```

> ### Challenge: violoin plot

> ### Challenge: box plot alternatives (bonus)


## Plotting time series data

Do the values change over years? within a year? seasonality?

### Have the species abundances changed over the years?

abundace = number of observations per species per year

"per" -> group_by()

```{r, purl=FALSE}
yearly_counts <- surveys_complete %>%
  group_by(year, species_id) %>% 
  summarise(n = n())
```

<!-- could also use count() -->

visualize as line plot with years on x and counts on y

```{r first-time-series, purl=FALSE}
ggplot(data = yearly_counts, mapping = aes(x = year, y = n)) +
     geom_line()
```

woa! what?
let's look at data for why

```{r}
head(yearly_counts, n=20)
```

need to group data by species then draw lines

```{r time-series-by-species, purl=FALSE}
ggplot(data = yearly_counts, mapping = aes(x = year, y = n, group = species_id)) +
    geom_line()
```

which is which? add colors per species

```{r time-series-with-colors, purl=FALSE}
ggplot(data = yearly_counts, mapping = aes(x = year, y = n, color = species_id)) +
    geom_line()
```

not really helpful... 

> ### Challenge: line plot

> ### Challenge: line plot with split-apply-combine - bonus

> ### Challenge: seasonality without facets (bonus bonus)


## Faceting

split one plot into multiple plots based on a category in the data

```{r first-facet, purl=FALSE}

ggplot(data = yearly_counts, mapping = aes(x = year, y = n)) +
    geom_line() +
    facet_wrap(.~species_id)
```

are there sex differences?

back to the data:

```{r, purl=FALSE}

countsByYearBySex <- surveys_complete %>%
  group_by(year, species_id, sex) %>% 
  summarise(n = n())

```

plot it

```{r facet-by-species-and-sex, purl=FALSE}
 ggplot(data = countsByYearBySex, mapping = aes(x = year, y = n, color = sex)) +
     geom_line() +
     facet_wrap(.~ species_id)
```


> ### Challenge - recap weight by year



Can skip: `facet_wrap` vs `facet_grid`

compare how the weights of males and females vs time, plot types


```{r average-weight-time-facet-sex-rows, purl=FALSE}
# One column, facet by rows
yearly_sex_weight <- surveys_complete %>%
    group_by(year, sex, species_id, plot_type) %>%
    summarize(avg_weight = mean(weight))

ggplot(data = yearly_sex_weight, 
       mapping = aes(x = year, y = avg_weight, color = species_id)) +
    geom_line() +
    facet_grid(sex ~ plot_type)
```

## Themes and Customization

build this up 1 line at a time or share code and walk through

```{r number-species-year-with-full-theme, purl=FALSE}

plotCountsByYearBySex <- ggplot(data = countsByYearBySex, mapping = aes(x = year, y = n, color = sex)) +
     geom_line() +
     facet_wrap(.~ species_id) +
    labs(title = "Observed species in time", # informative title and axis labels
        x = "Year of observation",
        y = "Number of individuals") +
    theme_bw() + #white background looks better printed+
    theme(panel.grid = element_blank(), # remove grid lines
          text = element_text(size = 16), # readability: increase font size
          axis.text.y = element_text(colour = "grey20", size = 12),
          axis.text.x = element_text(colour = "grey20", size = 12, 
                                     angle = 90, hjust = 0.5, vjust = 0.5) # rotate x axis, justify
          )

plotCountsByYearBySex
```


all themes: <http://docs.ggplot2.org/current/ggtheme.html>

popular: `theme_minimal()` , `theme_light()` 

`ggplot2` cheat sheet to improve the plot.

save plot e.g. for manuscript 

```{r}

dir.create(path = "fig_output")
ggsave(filename = "fig_output/countsByYearBySex.png",
       plot = plotCountsByYearBySex, 
       width = 15, height = 10,dpi = 300)

```


save theme to apply to next plot

```{r number-species-year-with-right-labels-xfont-orientation, purl=FALSE}
grey_theme <-theme(axis.text.y = element_text(colour = "grey20", size = 12),
                   axis.text.x = element_text(colour = "grey20", size = 12, 
                                              angle = 90, hjust = 0.5, vjust = 0.5),
                   text = element_text(size = 16))

ggplot(surveys_complete, aes(x = species_id, y = hindfoot_length)) +
    geom_boxplot() +
    grey_theme

```

> ### Challenge
> 
> With all of this information in hand, please take another five minutes to
> either improve one of the plots generated in this exercise or create a
> beautiful graph of your own. Use the RStudio [**`ggplot2`** cheat sheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)
> for inspiration. Here are some ideas:
>
> * See if you can change the thickness of the lines.
> * Can you find a way to change the name of the legend? What about its labels?
> * Try using a different color palette (see
>   http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/).

## Arranging plots

just demo this?

figure that contains multiple plots using different variables or even different data frames

**`gridExtra`** package allows us to combine separate ggplots into a single figure using `grid.arrange()`:

```{r install-gridextra, message=FALSE, purl=TRUE, eval=FALSE}
install.packages("gridExtra")
```

```{r gridarrange-example, message=FALSE, purl=FALSE, fig.width=10}
library(gridExtra)

spp_weight_boxplot <- ggplot(data = surveys_complete, 
                             mapping = aes(x = species_id, y = weight)) +
  geom_boxplot() +
  labs(x="Species", y= "Weight (g)") +
  scale_y_log10() + 
  grey_theme

spp_count_plot <- ggplot(data = yearly_counts, 
                         mapping = aes(x = year, y = n, color = species_id)) +
  geom_line() + 
  labs(x="Year", y= "Abundance") +
  grey_theme

grid.arrange(spp_weight_boxplot, spp_count_plot, ncol = 2, widths = c(4, 6))

combo_plot <- grid.arrange(spp_weight_boxplot, spp_count_plot, ncol = 2, widths = c(4, 6))

ggsave("fig_output/combo_plot_abun_weight.png", combo_plot, width = 10, dpi = 300)

```


Note: The parameters `width` and `height` also determine the font size in the saved plot.


```{r final-challenge, eval=FALSE, purl=TRUE, echo=FALSE}
### Final plotting challenge:
##  With all of this information in hand, please take another five
##  minutes to either improve one of the plots generated in this
##  exercise or create a beautiful graph of your own. Use the RStudio
##  ggplot2 cheat sheet for inspiration:
##  https://www.rstudio.com/wp-content/uploads/2015/08/ggplot2-cheatsheet.pdf
```

# Other

The `+` in the **`ggplot2`** package is particularly useful because it allows you
to modify existing `ggplot` objects. This means you can easily set up plot
templates and conveniently explore different types of plots, so the above
plot can also be generated with code like this:

```{r, first-ggplot-with-plus, eval=FALSE, purl=FALSE}
# Assign plot to a variable
surveys_plot <- ggplot(data = surveys_complete, 
                       mapping = aes(x = weight, y = hindfoot_length))

# Draw the plot
surveys_plot + 
    geom_point()
```

```{r, eval=FALSE, purl=TRUE, echo=FALSE, purl=FALSE}
# Create a ggplot and draw it
surveys_plot <- ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length))
surveys_plot + 
    geom_point()
```


------------

> ### Learning Objectives
>
> * Produce scatter plots, boxplots, and time series plots using ggplot.
> * Describe what faceting is and apply faceting in ggplot.
> * Modify the aesthetics of a ggplot plot (including axis labels and color).
> * Build complex and customized plots from data in a data frame.

--------------
