---
title: Data visualization with ggplot2
author: Thea Van Rossum
---

------------

> ### Learning Objectives
>
> * Produce scatter plots, boxplots, and time series plots using ggplot.
> * Describe what faceting is and apply faceting in ggplot.
> * Modify the aesthetics of a ggplot plot (including axis labels and color).
> * Build complex and customized plots from data in a data frame.

--------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,error=T)
```


 required: install.packages("gridExtra")

**`ggplot2`** is included in the **`tidyverse`** package.

```{r load-package, message=FALSE, purl=FALSE}
library(tidyverse)
```

load the data from prev. lesson

```{r load-data, purl=FALSE}
surveys_complete <- read_csv("/Users/rossum/Dropbox/teaching/Workshops/SCW-DCW/lessonsDC/R-ecology-lesson/data_output/surveys_complete.csv")
```

look at our data!

```{r}
str(surveys_complete)
```

# Plotting with **`ggplot2`**

build a plot using programming commands

- reproducible
- documented
- easy to change underlying data
- easy to change visualisation type
- reusable
- helper libraries


**`ggplot2`** functions like data in the 'long' format, i.e., a column for every dimension,
and a row for every observation. Well-structured data will save you lots of time
when making figures with **`ggplot2`**

# Is there a relationship between foot size and weight?

template:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

indicate the data to use

```{r, purl=FALSE}
ggplot(data = surveys_complete)
```

map variables to axes and colors using the aesthetic (`aes`) function

```{r, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length))
```

add 'geoms' – graphical representations of the data

many available

use the `+` operator

`+` goes at the end of the line (not begining of next line)

`+` vs `%>%` 

```{r first-ggplot, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

<!-- Challenge -->
<!-- Make a scatter (x-y) plot that shows weight measurements collected every year but only for data collected after 1990. Use both the `%>%` and `+` symbols in your answer -->
<!-- hint: remember the function `filter()` -->

<!-- ```{r} -->
<!-- surveys_complete_post1990 <- surveys_complete %>% filter(year > 1990) -->
<!-- ggplot(data = surveys_complete_post1990, mapping = aes(x = year, y = weight)) + -->
<!--   geom_point() -->

<!-- ``` -->

data investigation time!

start by making a plot like we did 
then modify it to display more information *or* less information 
communitcate message more clearly

# Is this relationship the same for all genera?

are genera different here? -> add color

```{r color-by-genus-1, purl=FALSE}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length)) +
    geom_point(alpha = 0.4, aes(color = genus))
```

### Challenge - scatter plot

### Challenge - overplotting - bonus

### Challenge: hex plot - optional  


# Boxplot

## do the genera have different weights?

visualize the distribution of weight within each genus

all the built-in geoms: https://ggplot2.tidyverse.org/reference/

use pipe for data

```{r boxplot, purl=FALSE}
surveys_complete %>% 
  ggplot(mapping = aes(x = genus, y = weight)) +
    geom_boxplot()
```

hide less: show points

```{r boxplot-with-points, purl=FALSE}
surveys_complete %>% 
ggplot(mapping = aes(x = genus, y = weight)) +
    geom_boxplot() +
    geom_jitter(height = 0, alpha = 0.3)
```

layer order

```{r boxplot-with-points-layer-order, purl=FALSE}
surveys_complete %>% 
ggplot(mapping = aes(x = genus, y = weight)) +
    geom_jitter(height = 0, alpha = 0.3)+
    geom_boxplot() 
```

color apply to all layers -> put in top `ggplot(aes())`

```{r color-by-genus-2, purl=FALSE}
surveys_complete %>% 
ggplot(mapping = aes(x = genus, y = weight, color = genus)) +
    geom_jitter(height = 0, alpha = 0.3)+
    geom_boxplot() 
```

### Challenge: violoin plot

### Challenge: box plot alternatives (bonus)


# Plotting time series data - group/summarise + line plot

Do the values change over years? within a year? seasonality?

## Have the genus abundances changed over the years?

abundace = number of observations per genus per year

"per" -> group_by()

```{r, purl=FALSE}
yearly_counts <- surveys_complete %>%
  group_by(year, genus) %>% 
  summarise(countsPerGenusPerYear = n())

```

<!-- could also use count() -->

visualize as line plot with years on x and counts on y

```{r first-time-series, purl=FALSE}

ggplot(data = yearly_counts, mapping = aes(x = year, y = countsPerGenusPerYear)) +
     geom_line()
```

woa! what?
let's look at data for why

```{r}
head(yearly_counts, n=20)
```

need to group data by genus then draw lines

```{r time-series-by-genus, purl=FALSE}
ggplot(data = yearly_counts, 
       mapping = aes(x = year, y = countsPerGenusPerYear, group = genus)) +
    geom_line()
```

which is which? add colors per genus

```{r time-series-with-colors, purl=FALSE}
ggplot(data = yearly_counts, 
       mapping = aes(x = year, y = countsPerGenusPerYear, color = genus)) +
    geom_line()
```


### Challenge: line plot - median weight by month

### Bonus challenge: variability in trend



# Faceting

## improve visibility of previous genus/time plot

split one plot into multiple plots based on a category in the data

```{r first-facet, purl=FALSE}

ggplot(data = yearly_counts, 
       mapping = aes(x = year, y = countsPerGenusPerYear)) +
    geom_line() +
    facet_wrap(.~genus)
```

## Are there sex differences in the time trends?

back to the data:

```{r, purl=FALSE}

countsByYearBySex <- surveys_complete %>%
  group_by(year, genus, sex) %>% 
  summarise(countsPerGenusPerYearPerSex = n())

```

plot it

```{r facet-by-genus-and-sex, purl=FALSE}
 ggplot(data = countsByYearBySex, 
        mapping = aes(x = year, y = countsPerGenusPerYearPerSex, color = sex)) +
     geom_line() +
     facet_wrap(.~ genus)
```


### Challenge - sex diff by species?

### Challenge - weight by year - recap

### Bonus challenge: investigate species "RF"





Can skip: `facet_wrap` vs `facet_grid`

compare how the weights of males and females vs time, plot types

```{r average-weight-time-facet-sex-rows, purl=FALSE}
# One column, facet by rows
yearly_sex_weight <- surveys_complete %>%
    group_by(year, sex, genus, plot_type) %>%
    summarize(avg_weight = mean(weight))

ggplot(data = yearly_sex_weight, 
       mapping = aes(x = year, y = avg_weight, color = genus)) +
    geom_line() +
    facet_grid(sex ~ plot_type)
```


# Factors - months

Does the number of animals observed in the genera "Onychomys" and "Perognathus" change from month to month across the years studied?

```{r}

countsPerMonthPerYear <- surveys_complete %>%
  filter(genus %in% c("Onychomys", "Perognathus")) %>% 
  group_by(year, month, genus) %>% 
  summarise(observationsPerGenusPerMonthPerYear = n())

countsPerMonthPerYear %>% 
  ggplot(aes(x=month,y=observationsPerGenusPerMonthPerYear,group=month))+
  geom_boxplot() +
  facet_wrap(. ~ genus)
```

get abbrev names of months 

```{r}

month.abb
month.abb[c(1,5,3)]
countsPerMonthPerYear$month %>% head()
month.abb[countsPerMonthPerYear$month] %>% head()

```

use this on our data

```{r}

countsPerMonthPerYear %>% 
  mutate(monthNames = month.abb[month]) %>%  #<------------------ ADDED THIS
  ggplot(aes(x=monthNames,y=observationsPerGenusPerMonthPerYear,group=monthNames))+
  geom_boxplot() +
  facet_wrap(. ~ genus)

```

need factors to get the right order


Factors are very useful; contribute to making R particularly well suited to working with data

- Factors represent categorical data. 
- ordered or unordered, default = alpha
- stored as integers associated with labels
- factors can only contain a pre-defined set of values, known as *levels*


```{r, purl=TRUE}
sex <- factor(c("male", "female", "female", "male"))
```

R will assign 
- `1` to the level `"female"`  (alphabetical order)
- `2` to the level `"male"` 

`levels()` 

`nlevels()`

```{r, purl=FALSE}
levels(sex)
nlevels(sex)
```

Sometimes, the order of the factors does not matter, other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high"),
it improves your visualization, or it is required by a particular type of
analysis. Here, one way to reorder our levels in the `sex` vector would be:

```{r, results=TRUE, purl=FALSE}
sex # current order
sex <- factor(sex, levels = c("male", "female"))
sex # after re-ordering
```

In R's memory, these factors are represented by integers (1, 2, 3), but are more
informative than integers because factors are self describing: `"female"`,
`"male"` is more descriptive than `1`, `2`. Which one is "male"?  You wouldn't
be able to tell just from the integer data. Factors, on the other hand, have
this information built in. It is particularly helpful when there are many levels
(like the species names in our example dataset).

### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl=FALSE}
as.character(sex)
```

`as.numeric()` BAD - gives levels

`as.character()` then `as.numeric()`

Another method is to use the `levels()` function. Compare:

```{r, purl=TRUE}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct)               # Wrong! And there is no warning...
as.numeric(as.character(year_fct)) # Works...
```

back to our example 

get the right order for months

```{r}

countsPerMonthPerYearToPlot <- countsPerMonthPerYear %>% 
  mutate(monthNames = month.abb[month]) %>% 
  mutate(monthNames = factor(monthNames,ordered = T,levels = month.abb))  #<------------------ ADDED THIS

glimpse(countsPerMonthPerYearToPlot)

countsPerMonthPerYearToPlot %>% 
  ggplot(aes(x=monthNames,y=observationsPerGenusPerMonthPerYear,group=monthNames))+
  geom_boxplot() +
  facet_wrap(. ~ genus)
  
```

how to detect and avoid factors
- glimpse() or str()
- use read_csv not read.csv
- strings_as_factors=F

## Challenge - reorder genera in plot

Use a factor to reorder the genera in the plot we just made so that "Perognathus" comes before "Onychomys"

```{r}
countsPerMonthPerYearToPlot %>% 
  mutate(genus = factor(genus,ordered = T,levels = c("Perognathus", "Onychomys"))) %>% 
  ggplot(aes(x=monthNames,y=observationsPerGenusPerMonthPerYear,group=monthNames))+
    geom_boxplot() +
    facet_wrap(. ~ genus)
```



# Themes and Customization

lets use that species plot and fix it up

build this up 1 line at a time or share code and walk through

```{r number-species-year-with-full-theme, purl=FALSE}

plotCountsByYearBySexBySpecies <- surveys_complete %>%
    group_by(year, sex, species_id) %>%
    summarize(n = n()) %>% 
  ggplot(mapping = aes(x = year, y = n, color = sex)) +
     geom_line() +
     facet_wrap(.~ species_id) +
    labs(title = "Observed genera in time", # informative title and axis labels
        x = "Year of observation",
        y = "Number of individuals") +
    theme_bw() + #white background looks better printed+
    theme(panel.grid = element_blank(), # remove grid lines
          text = element_text(size = 16), # readability: increase font size
          axis.text.y = element_text(colour = "grey20", size = 12),
          axis.text.x = element_text(colour = "grey20", size = 12, 
                                     angle = 90, hjust = 0.5, vjust = 0.5) # rotate x axis, justify
          )

plotCountsByYearBySexBySpecies
```


all themes: <http://docs.ggplot2.org/current/ggtheme.html>

popular: `theme_minimal()` , `theme_light()` 

`ggplot2` cheat sheet to improve the plot.

# Saving plots

save plot e.g. for manuscript 

```{r}

dir.create(path = "fig_output")
ggsave(filename = "fig_output/countsByYearBySex.png",
       plot = plotCountsByYearBySexBySpecies, 
       width = 15, height = 10,dpi = 300)

```

# Resusing custom themes

save theme to apply to next plot

```{r number-genus-year-with-right-labels-xfont-orientation, purl=FALSE}
grey_theme <-theme(axis.text.y = element_text(colour = "grey20", size = 12),
                   axis.text.x = element_text(colour = "grey20", size = 12, 
                                              angle = 90, hjust = 0.5, vjust = 0.5),
                   text = element_text(size = 16))

ggplot(surveys_complete, aes(x = genus, y = hindfoot_length)) +
    geom_boxplot() +
    grey_theme

```

### Challenge -- free style


## Arranging plots

just demo this?

figure that contains multiple plots using different variables or even different data frames

**`gridExtra`** package allows us to combine separate ggplots into a single figure using `grid.arrange()`:


```{r gridarrange-example, purl=FALSE, fig.width=10}
library(gridExtra)

spp_weight_boxplot <- ggplot(data = surveys_complete, 
                             mapping = aes(x = genus, y = weight)) +
  geom_boxplot() +
  labs(x="Genus", y= "Weight (g)") +
  scale_y_log10() + 
  grey_theme

spp_count_plot <- ggplot(data = yearly_counts, 
                         mapping = aes(x = year, y = n, color = genus)) +
  geom_line() + 
  labs(x="Year", y= "Abundance") +
  grey_theme

grid.arrange(spp_weight_boxplot, spp_count_plot, ncol = 2, widths = c(4, 6))

combo_plot <- grid.arrange(spp_weight_boxplot, spp_count_plot, ncol = 2, widths = c(4, 6))

ggsave("fig_output/combo_plot_abun_weight.png", combo_plot, width = 10, dpi = 300)

```


Note: The parameters `width` and `height` also determine the font size in the saved plot.


```{r final-challenge, purl=TRUE, echo=FALSE}
### Final plotting challenge:
##  With all of this information in hand, please take another five
##  minutes to either improve one of the plots generated in this
##  exercise or create a beautiful graph of your own. Use the RStudio
##  ggplot2 cheat sheet for inspiration:
##  https://www.rstudio.com/wp-content/uploads/2015/08/ggplot2-cheatsheet.pdf
```


Resources:

- built-in themes https://ggplot2.tidyverse.org/reference/ggtheme.html
- ggplot extensions: http://www.ggplot2-exts.org/gallery/
- gallery of plots - click on plot to see code https://www.r-graph-gallery.com/portfolio/ggplot2-package/
- more plots with code: http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html#1.%20Correlation 
- material in general here is good: http://r-statistics.co/ 
- common plotting customisations http://r-statistics.co/ggplot2-cheatsheet.html 



# Other

The `+` in the **`ggplot2`** package is particularly useful because it allows you
to modify existing `ggplot` objects. This means you can easily set up plot
templates and conveniently explore different types of plots, so the above
plot can also be generated with code like this:

```{r, first-ggplot-with-plus, purl=FALSE}
# Assign plot to a variable
surveys_plot <- ggplot(data = surveys_complete, 
                       mapping = aes(x = weight, y = hindfoot_length))

# Draw the plot
surveys_plot + 
    geom_point()
```

```{r, purl=TRUE, echo=FALSE, purl=FALSE}
# Create a ggplot and draw it
surveys_plot <- ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length))
surveys_plot + 
    geom_point()
```


------------

> ### Learning Objectives
>
> * Produce scatter plots, boxplots, and time series plots using ggplot.
> * Describe what faceting is and apply faceting in ggplot.
> * Modify the aesthetics of a ggplot plot (including axis labels and color).
> * Build complex and customized plots from data in a data frame.

--------------




## Factors can improve visualisation

```{r}

```

```{r}
head(surveys_complete)
```


* Using your `yearly_diversity` table, can you recreate the plot below, ensuring
the order of the `plot_type` facets is as shown?

```{r}
# Make a vector with treatments in desired order
## this is optional, but makes it easier to read the code
plot_type_order <- c("Control",
                     "Short-term Krat Exclosure", "Long-term Krat Exclosure",
                     "Rodent Exclosure", "Spectab exclosure")

# Factorise the `plot_type` variable with custom levels before plotting
yearly_diversity %>%
  mutate(plot_type = factor(plot_type, levels = plot_type_order)) %>%
  ggplot(aes(year, n_species)) +
  geom_line(aes(colour = sex)) +
  labs(colour = "Plot Type", x = "Year", y = "# species") +
  facet_wrap( ~ plot_type)
```

